---
title: "Geospatial workshop"
author: "Lydia Walton"
date: "22/03/2023"
output: html_document
---
# Intro to geospatial data
Geospatial vector data consist of geometries defined based on geographic coordinates.https://ourcodingclub.github.io/tutorials/spatial-vector-sf/ 
```{r}
#Simple features is a set of standards for geospatial data. 

# Load libraries
library(dplyr) # data wrangling
library(tidyr) # data wrangling
library(ggplot2) # data visualisation
library(sf) # simple features - geospatial geometries
library(osmdata) # obtaining OpenStreetMap vector data
library(units) # working with units
library(mapview) # interactive geometry viewing
library(ggmap) # downloading raster maps from a variety of sources
library(ggspatial) # map backgrounds and annotations for ggplot
library(tmap) # static/interactive map library with ggplot-like syntax

```

# Edinburgh data
```{r}
# Get the polygon for Edinburgh
city_polygon <- getbb("City of Edinburgh",
                      featuretype = "settlement",
                      format_out = "polygon")

# Get the rectangular bounding box
city_rect <- getbb("City of Edinburgh", featuretype = "settlement")

#Pull out green spaces
greensp <- opq(bbox = city_polygon) %>% 
  add_osm_feature(key = "leisure", value = c("park", "nature_reserve", "golf_course")) %>% 
  osmdata_sf() %>% 
  trim_osmdata(city_polygon)

greensp
  
glimpse(greensp$osm_polygons)

#multi-polygon binding
greensp_sf <- bind_rows(st_cast(greensp$osm_polygons, "MULTIPOLYGON",
                                greensp$osm_multipolygons)) %>% 
  select(name, osm_id, leisure)

plot(greensp_sf["leisure"])

head(greensp_sf)
unique(greensp_sf$leisure)

#remove NAs
newgreensp_sf <- greensp_sf %>% 
  filter(is.na(leisure) == FALSE) %>% 
  rename(greensp_type = leisure) %>% 
  st_make_valid()

#Save this data to upload it later (saves as a geopackage ".gpkg")
st_write(newgreensp_sf,
         dsn = "greenspaces_Edi_OSM.gpkg", # file path
         layer="greenspaces", # layer name
         layer_options = c(paste0("DESCRIPTION=Contains spatial multipolygons for parks, ",
                                  "nature reserves and golf courses in Edinburgh, Scotland. ",
                                  "Copyright OpenStreetMap constibutors. ODbL ",
                                  "https://www.openstreetmap.org/copyright")),
         # add layer description
         delete_dsn = TRUE
         # to delete the whole file first, because sometimes, we can just
         # overwrite or append one layer to an already existing file.
         # If the file doesn't exist, this will return a friendly warning.
         )

#Read in the data
gsp <- st_read(dsn = "greenspaces_Edi_OSM.gpkg", layer = "greenspaces")
gsp
```

# Coordinate Reference Systems (CRS)

Coordinate reference systems relate vector geometries to the Earth’s surface, and using the right CRS can be very important to execute our operations successfully. There are two main types of CRS: geographic and projected.

3D object to a flat map --> projections are how we project 3D globe onto 2D surface
We will set our projections in R so everything can stack properly and we don't have any mismatches

WGS 84 is generally a good projection for North America 
```{r}
#Look at the CRS of our sf object
st_crs(greensp_sf)

#Transform projection to British National Grid (OSGB 1936)
greensp_sf <- st_transform(greensp_sf, 27700)
st_crs(greensp_sf)


```

# Spatial Operations
```{r}
#Create new column within our df
greensp_sf <- mutate(greensp_sf, area = st_area(greensp_sf))

head(greensp_sf$area) #Area for each spatial polygon object (in m^2)

#Set units by hand
greensp_sf <- greensp_sf %>% 
  mutate(area_ha = set_units(area, "ha")) #Change units to ha

#mapview can create an interactive map
mapview(greensp_sf)

#Filter out areas that are too small, etc.
greensp_sf <- filter(greensp_sf, as.numeric(area_ha)>=2)

#Can split into different lists - golf course, nature reserve, park
gsp_list <- greensp_sf %>% 
  split(.$leisure)

#Cut out parts of parks that are covered by nature reserves or golf course (remove the overlap)
gsp_list$park <- st_difference(gsp_list$park, st_union(gsp_list$nature_reserve))
gsp_list$park <- st_difference(gsp_list$park, st_union(gsp_list$golf_course))

gsp <- bind_rows(gsp_list) %>% 
 mutate(area = st_area(.), 
        area_ha = set_units(area, "ha")) %>% 
  filter(as.numeric(area_ha)>=2)

mapview(gsp)

```

# Draw Maps
```{r}
#Create new df with mutated levels of green spaces (Replace the way things are spelt)
greensp_sf_forplot <- mutate(gsp, leisure = as.factor(factor(leisure, levels = c("golf_course", "nature_reserve", "park"), labels = c("Golf Course", "Nature Reserve", "Park"))))

#First make a Stamen map 
stamen_raster <- get_stamenmap(city_rect, zoom = 12)

# Plot map with ggplot
(edi_greenspaces_map <-
  ggplot(data = greensp_sf_forplot) +
    inset_ggmap(stamen_raster) + # add ggmap background
    geom_sf(aes(fill = leisure)) + # add sf shapes, coloured by greensp_type
    coord_sf(crs = st_crs(4326), expand = FALSE) +
    # change the CRS of the sf back to WGS84 to match the ggmap raster
    scale_fill_manual(values = c("#44AA99", "#117733", "#AA4499")) +
    # add custom colours from Tol palette (colourblind-friendly)
    labs(title = "Green spaces in Edinburgh, Scotland",
         subtitle = "Parks, nature reserves and golf courses > 2 ha\n",
         caption = paste0("Map tiles by Stamen Design (stamen.com), CC BY 3.0. ",
                         "http://creativecommons.org/licenses/by/3.0\n",
                         "Map data © OpenStreetMap contributors, ODbL. ",
                         "http://www.openstreetmap.org/copyright")) +
    # add various labels
    annotation_scale(location = "bl") + # ggspatial scale on bottom left
    annotation_north_arrow(location = "tr") + # ggspatial arrow on top right
    theme_void() + # get rid of axis ticks, titles
    theme(legend.title = element_blank(),
          legend.position = c(.98, .02),
          legend.justification = c("right", "bottom"),
          legend.box.just = "right",
          legend.box.background = element_rect(fill = "white", colour = "gray"),
          legend.margin = margin(6, 6, 6, 6),
          # move legend to bottom right and customise
          plot.margin = margin(12,12,12,12))
          # add margin around plot
)

```

# Make an interactive map using tmap
```{r}
# Plot interactively with tmap
tmap_mode("view") # interactive mode

(edi_greenspace_tmap <-
  tm_basemap("Stamen.Terrain") + # add Stamen Terrain basemap
  tm_shape(greensp_sf_forplot) + # add the sf
  tm_sf(col = "leisure", # colour by green space type
        title = "", # no legend title
        palette = c("#44AA99", "#117733", "#AA4499"), # custom fill colours
        popup.vars = c("Area  " = "area_ha"), # customise popup to show area
        popup.format = list(digits=1)) + # limit area to 1 decimal digit
  tm_scale_bar() # add scale bar
)
```

